---
title: "Convert GO terms annotation parsed with Python file into GAF format"
output: html_notebook
---

# Convert the GO terms annotation parsed with Python file into GAF format 


```{r}


#Test if BioManager is installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install()
}

# load pacman package manager
if (!require(pacman)) {
  install.packages("pacman")
  library(pacman)
}

p_load(tidyverse)
p_load(plotly)
p_load(vroom)
p_load(writexl)
p_load(ggplot2)
p_load(GO.db)
p_load(clusterProfiler)

p_load(magrittr)
p_load(xml2)
p_load(rvest)
p_load(ProteomeRiver)

```

```{r}
base_dir <-  here::here() 
data_dir <- file.path( base_dir, "Data")
results_dir <- file.path(base_dir, "Results")
source_dir <- file.path(base_dir, "Source")


```

```{r}
go_annot <- vroom::vroom( file.path( results_dir, "UniProt", "go_terms_table_python.tab")  ) 

# UniProt entries tab separated format to obtain gene symbols
uniprot_tbl <- vroom::vroom( file.path( data_dir, "UniProt", "data.tab"))  %>%
  dplyr::mutate(gene_symbol = str_split(`Gene names`, " ") %>% purrr::map_chr( 1)) %>%
  dplyr::relocate("gene_symbol", .before="Entry name")
```
# http://geneontology.org/docs/go-annotation-file-gaf-format-2.2/
```{r}
go_annot
```

# GAF file required headers
```{r}


```

## Data fields required by the GAF format 

Column	Content	Required?	Cardinality	Example	 
1	DB	required	1	UniProtKB	 
2	DB Object ID	required	1	P12345	 
3	DB Object Symbol	required	1	PHO3	 
4	Qualifier	required	1 or 2	NOT|involved_in	 
5	GO ID	required	1	GO:0003993	 
6	DB:Reference (|DB:Reference)	required	1 or greater	PMID:2676709	 
7	Evidence Code	required	1	IMP	 
8	With (or) From	optional	0 or greater	GO:0000346	 
9	Aspect	required	1	F	 
10	DB Object Name	optional	0 or 1	Toll-like receptor 4	 
11	DB Object Synonym (|Synonym)	optional	0 or greater	hToll	Tollbooth
12	DB Object Type	required	1	protein	 
13	Taxon(|taxon)	required	1 or 2	taxon:9606	 
14	Date	required	1	20090118	 
15	Assigned By	required	1	SGD	 
16	Annotation Extension	optional	0 or greater	part_of(CL:0000576)	 
17	Gene Product Form ID	optional	0 or 1	UniProtKB:P12345-2	 


```{r}
gaf_tbl <- go_annot %>%
  left_join( uniprot_tbl %>%
               dplyr::select(Entry, gene_symbol), by =c("uniprot_acc" = "Entry")) %>%
  dplyr::mutate( DB = "UniProtKB") %>%
  dplyr::relocate( DB, .before="uniprot_acc") %>% 
  dplyr::rename( 	DB_Object_ID = "uniprot_acc" ) %>%
  dplyr::relocate( gene_symbol, .after="DB_Object_ID") %>% 
  dplyr::rename( 	DB_Object_Symbol = "gene_symbol" ) %>%
  dplyr::mutate(  Qualifier = "involved_in" ) %>%
  dplyr::relocate( Qualifier, .after="DB_Object_Symbol") %>%
  dplyr::rename( GO_ID = "go_id")  %>%
  dplyr::mutate( DB_Reference = paste0("UniProtKB:", DB_Object_ID) ) %>%
  dplyr::relocate(DB_Reference, .after="GO_ID" ) %>%
  dplyr::rename( Evidence_Code = "go_evidence") %>%
  dplyr::relocate( Evidence_Code, .after="DB_Reference") %>%
  dplyr::mutate( With_or_From = "") %>%
  dplyr::relocate( With_or_From, .after="Evidence_Code") %>%
  dplyr::rename( Aspect = "go_type")  %>%
  dplyr::rename( DB_Object_Name = "go_term") %>%
  dplyr::mutate( DB_Object_Synonym = "" ) %>%
  dplyr::relocate( DB_Object_Synonym, .after="DB_Object_Name") %>%
  dplyr::mutate( DB_Object_Type = "protein") %>%
  dplyr::relocate( DB_Object_Type, .after="DB_Object_Synonym") %>%
  dplyr::mutate( Taxon = "taxon:9606") %>%
  dplyr::relocate( Taxon, .after="DB_Object_Type") %>%
  dplyr::mutate( Date  =  format(Sys.Date(), "%Y%m%d") ) %>%
  dplyr::relocate( Date, .after="Taxon") %>%
  dplyr::rename(Assigned_By = "go_project" )  %>%
  dplyr::relocate( Assigned_By, .after="Date") %>%
  dplyr::mutate( Annotation_Extension = "") %>%
  dplyr::mutate( Gene_Product_Form_ID = DB_Reference)

gaf_tbl


colnames( gaf_tbl)
```
## Create GAF file 
```{r}
sink( file.path(results_dir, "UniProt", "go_annotation_header.gaf" ))
cat("!gaf-version: 2.2\n")
cat("generated-by: UniProt\n")
cat("date-generated: 2022-04-22\n")
sink()

vroom::vroom_write(  gaf_tbl, col_names = FALSE, path=file.path(results_dir, "UniProt", "go_annotation_body.gaf") ) 

system(  paste( "cat",
  file.path(results_dir, "UniProt", "go_annotation_header.gaf" ),
  file.path(results_dir, "UniProt", "go_annotation_body.gaf"),
  ">",
  file.path(results_dir, "UniProt", "go_annotation.gaf") ) ) 

```


## 
```{r}

go_slim_tbl <-  vroom::vroom(file.path( results_dir, "UniProt", "go_annotation.slim.gaf"),   comment = "!",
                             col_names = c("DB", "DB_Object_ID", "DB_Object_Symbol", "Qualifier", "GO_ID", 
                                           "DB_Reference", "Evidence_Code", "With_or_From",  "Aspect", 
                                           "DB_Object_Name", "DB_Object_Synonym", "DB_Object_Type", "Taxon", 
                                           "Date", "Assigned_By", "Annotation_Extension", "Gene_Product_Form_ID" )) 


permitted_go_terms <- go_slim_tbl %>% 
  group_by( GO_ID) %>%
  dplyr::summarize( counts =n()) %>%
  ungroup() %>%
  arrange(desc(counts)) %>%
  dplyr::filter( counts <= 2000 & counts >= 10 )  %>%
  dplyr::mutate( go_term = purrr::map_chr(  GO_ID, ~Term(.)))

# Number of proteins annotated (13,460)
go_slim_tbl %>%
  dplyr::inner_join(permitted_go_terms, by ="GO_ID") %>%
  distinct(DB_Object_ID) %>%
  nrow()


go_slim_tbl_filtered <- go_slim_tbl %>%
  dplyr::inner_join(permitted_go_terms, by ="GO_ID") %>%
  dplyr::select(DB_Object_ID, GO_ID, Aspect, DB_Object_Name, Evidence_Code )

Term("GO:0042060")


vroom::vroom_write( go_slim_tbl_filtered, path=file.path(results_dir, "GOSubset", "go_slim_tbl_filtered.tab") )

```



